---
title: "UCSBees Analysis 2020"
author: "Charlie Thrift"
date: "1/31/2020"
output: html_document
---
##add mellitidae family??
##change to be able to see significance of small values on plot 1
##change the list on first plot, so that the legend isn't genus, its familyGenus
##unwrapped can show more diversity
##show distribution of halictidae on santa cruz island based on date
##can add dayof year to density
##see if we can combine the halictidae data with other halictidae data to see if the grouping of males and females is consistant, but also to see if there is more of a signal
##see if there are differences in site genera time, like one site eucera is may, other site is june
##going forward, thinking about presentations, thinking about what the driving questions are so that it can be presented to the audience

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages
Here, I will load necessary packages.
```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
```

# Load data
Here, I will load necessary data. This is all of the hymenoptera data 
from CCBER gbif site
```{r}
data <- read.csv(file = "data/hymenopteradata04mar2020.csv")
```

# Filter data
Here, I will pull only the UCSBees project data from all of the Hymenoptera.
```{r}
data <- filter(data, str_detect(recordedBy, pattern = "UCSBees"))
```

# Assign sites to the data
Form 8 different standard sites for the collected specimen, based on Latitude and 
Longitude from the data.
```{r}
#for North Campus Open Space 1 thru 8
data$site[data$decimalLatitude == 34.4215 & 
               data$decimalLongitude == -119.8827]<- "North Campus Open Space"
data$site[data$decimalLatitude == 34.4213 & 
               data$decimalLongitude == -119.8844]<- "North Campus Open Space"
data$site[data$decimalLatitude == 34.4183 & 
               data$decimalLongitude == -119.8776]<- "North Campus Open Space"
data$site[data$decimalLatitude == 34.4191 & 
               data$decimalLongitude == -119.8766]<- "North Campus Open Space"
data$site[data$decimalLatitude == 34.4189 & 
               data$decimalLongitude == -119.8799]<- "North Campus Open Space"
data$site[data$decimalLatitude == 34.4189 & 
               data$decimalLongitude == -119.8827]<- "North Campus Open Space"
data$site[data$decimalLatitude == 34.4197 & 
               data$decimalLongitude == -119.8818]<- "North Campus Open Space"
data$site[data$decimalLatitude == 34.4188 & 
               data$decimalLongitude == -119.8807]<- "North Campus Open Space"
#for Coal Oil Point Reserve 1 thru 6
data$site[data$decimalLatitude == 34.41238 & 
               data$decimalLongitude == -119.88110]<- "Coal Oil Point Reserve"
data$site[data$decimalLatitude == 34.41146 & 
               data$decimalLongitude == -119.88210]<- "Coal Oil Point Reserve"
data$site[data$decimalLatitude == 34.41276 & 
               data$decimalLongitude == -119.88231]<- "Coal Oil Point Reserve"
data$site[data$decimalLatitude == 34.41491 & 
               data$decimalLongitude == -119.88020]<- "Coal Oil Point Reserve"
data$site[data$decimalLatitude == 34.41733 & 
               data$decimalLongitude == -119.87962]<- "Coal Oil Point Reserve"
data$site[data$decimalLatitude == 34.40870 & 
               data$decimalLongitude == -119.87859]<- "Coal Oil Point Reserve"
#for North Parcel sites 1 thru 3
data$site[data$decimalLatitude == 34.4226 & 
               data$decimalLongitude == -119.8815]<- "North Parcel"
data$site[data$decimalLatitude == 34.4235 & 
               data$decimalLongitude == -119.8815]<- "North Parcel"
data$site[data$decimalLatitude == 34.4224 & 
               data$decimalLongitude == -119.8794]<- "North Parcel"
#for Santa Cruz Island sites 1 thru 7
data$site[data$decimalLatitude == 34.001056 & 
               data$decimalLongitude == -119.686972]<- "Santa Cruz Island"
data$site[data$decimalLatitude == 33.991972 & 
               data$decimalLongitude == -119.689278]<- "Santa Cruz Island"
data$site[data$decimalLatitude == 33.997278 & 
               data$decimalLongitude == -119.714056]<- "Santa Cruz Island"
data$site[data$decimalLatitude == 33.991056 & 
               data$decimalLongitude == -119.719278]<- "Santa Cruz Island"
data$site[data$decimalLatitude == 34.018889 & 
               data$decimalLongitude == -119.871111]<- "Santa Cruz Island"
data$site[data$decimalLatitude == 34.0225 & 
               data$decimalLongitude == -119.876667]<- "Santa Cruz Island"
data$site[data$decimalLatitude == 34.0175 & 
               data$decimalLongitude == -119.682222]<- "Santa Cruz Island"
#for CCBER Greenhouse sites 1 thru 2
data$site[data$decimalLatitude == 34.420935 & 
               data$decimalLongitude == -119.858054 ]<- "CCBER Greenhouse"
data$site[data$decimalLatitude == 34.420828 & 
               data$decimalLongitude == -119.858029 ]<- "CCBER Greenhouse"
#for UCSB Lagoon sites 1 thru 5
data$site[data$decimalLatitude == 34.407448 & 
               data$decimalLongitude == -119.849818 ]<- "UCSB Lagoon"
data$site[data$decimalLatitude == 34.40677 & 
               data$decimalLongitude == -119.84716 ]<- "UCSB Lagoon"
data$site[data$decimalLatitude == 34.407967 & 
               data$decimalLongitude == -119.848322 ]<- "UCSB Lagoon"
data$site[data$decimalLatitude == 34.408304 & 
               data$decimalLongitude == -119.848728 ]<- "UCSB Lagoon"
data$site[data$decimalLatitude == 34.40928 & 
               data$decimalLongitude == -119.84892 ]<- "UCSB Lagoon"
#for Santa Clara River sites 1 thru 8
data$site[data$decimalLatitude == 34.363911 & 
               data$decimalLongitude == -118.999505 ]<- "Santa Clara River"
data$site[data$decimalLatitude == 34.363902 & 
               data$decimalLongitude == -118.993442 ]<- "Santa Clara River"
data$site[data$decimalLatitude == 34.389062 & 
               data$decimalLongitude == -118.889657  ]<- "Santa Clara River"
data$site[data$decimalLatitude == 34.388092 & 
               data$decimalLongitude == -118.889530  ]<- "Santa Clara River"
data$site[data$decimalLatitude == 34.389105 & 
               data$decimalLongitude == -118.885348 ]<- "Santa Clara River"
data$site[data$decimalLatitude == 34.355015 & 
               data$decimalLongitude == -119.014507  ]<- "Santa Clara River"
data$site[data$decimalLatitude == 34.390363 & 
               data$decimalLongitude == -118.89511 ]<- "Santa Clara River"
data$site[data$decimalLatitude == 34.387065 & 
               data$decimalLongitude == -118.891438  ]<- "Santa Clara River"
#for Carpinteria Salt Marsh sites 1 thru 4
data$site[data$decimalLatitude == 34.0175 & 
               data$decimalLongitude == -119.682222 ]<- "Carpinteria Salt Marsh Reserve"
data$site[data$decimalLatitude == 34.40114 & 
               data$decimalLongitude == -119.53040  ]<- "Carpinteria Salt Marsh Reserve"
data$site[data$decimalLatitude == 34.39822 & 
               data$decimalLongitude == -119.53040  ]<- "Carpinteria Salt Marsh Reserve"
data$site[data$decimalLatitude == 34.39715 & 
               data$decimalLongitude == -119.52728 ]<- "Carpinteria Salt Marsh Reserve"
```

# Subset the data and remove the non-bees
Here, I will clean up the data by only using specific columns, and omitting others. Also, the non-bee bycatch from the UCSBees project are removed here.
```{r}
data <- data[,c("catalogNumber", "family", "genus", "species",
                   "decimalLatitude", "decimalLongitude", "eventDate",
                   "day", "month", "year", "identifiedBy", "sex", "startDayOfYear", "site", "fieldNumber")]
data <- na.omit(data, cols="site") #remove rows that have NA as the site (because
#if they don't have a site then their lat/long is not one of the standard collection sites)
data <- data[data$family %in% c("Andrenidae", "Apidae", "Colletidae", 
                                "Halictidae", "Megachilidae"), ]
```

# Plot the data
Plot count per family at each individual site, with genus represented by color.
```{r}
count_persite_bygenus_wrap <- ggplot(data, aes(x=family)) +
  geom_bar(aes(fill=genus)) + facet_wrap(~site) + 
  theme(axis.text.x = element_text(angle = 90))
count_persite_bygenus_wrap
```

# Plot 2: Families of bees over time
Plot all the bees over one year. Scatterplot of counts of bees over time, with family represented by color.
```{r}
#count_overtime_byfamily <- ggplot(data, aes(x=month, y=number)) + 
  #geom_point(aes(fill=family)) 
#count_overtime_byfamily
```

# Plot 3: Density of the bees through the year
Plot density of all bees through the year, and density through year per site.
```{r}
density_plot_all <- ggplot(data, aes(x=month)) + geom_density() 
density_plot_all
density_plot_all_byfamily <- ggplot(data, aes(x=month)) + geom_density(aes(color=family)) 
density_plot_all_byfamily
density_plot_all_byfamily_dayofyear <- ggplot(data, aes(x=startDayOfYear)) + geom_density(aes(color=family)) 
density_plot_all_byfamily_dayofyear
density_plot_facet <- ggplot(data, aes(x=month)) + geom_density() + facet_wrap(~site)
density_plot_facet
density_plot_facet_byfamily <- ggplot(data, aes(x=month)) + geom_density(aes(color=family)) +  facet_wrap(~site)
density_plot_facet_byfamily
```

# Make a new column for month name, and then make monthName plot chronologically
Use the "month" column value (1,2,3,...,12) to make "MonthName" column (Jan, Feb, Mar,...,Dec). Then, make monthName plot chronologically instead of alphabetically.
```{r}
#Make new column for the name of the month based on number
data$monthName[data$month == 1]<- "Jan"
data$monthName[data$month == 2]<- "Feb"
data$monthName[data$month == 3]<- "Mar"
data$monthName[data$month == 4]<- "Apr"
data$monthName[data$month == 5]<- "May"
data$monthName[data$month == 6]<- "Jun"
data$monthName[data$month == 7]<- "Jul"
data$monthName[data$month == 8]<- "Aug"
data$monthName[data$month == 9]<- "Sep"
data$monthName[data$month == 10]<- "Oct"
data$monthName[data$month == 11]<- "Nov"
data$monthName[data$month == 12]<- "Dec"

#Make this new monthName column plot chronologically.
data$monthName = factor(data$monthName, levels = month.abb)
```

# Clean the field numbers
NCOS has both NCOS_7a and NCOS_07a, for example
```{r}
data$fieldNumber[data$fieldNumber == "NCOS_7a"]<- "NCOS_07a"
data$fieldNumber[data$fieldNumber == "NCOS_7b"]<- "NCOS_07b"
```


# Check which months are missing data
Scatterplot site by month. Each site should have a point for each month. If it doesn't, then that month's worth of data still needs to be processed. This does not account for all the data from a month. For example, if only half of the UCSB Lagoon bees were processed in January, it looks like they all have been.
```{r}
check_missingdata_sitebymonth <- ggplot(data, aes(x=monthName, y=site)) + geom_point()
check_missingdata_sitebymonth
check_missingdata_trapbymonth_wrapsite <- ggplot(data, aes(x=monthName, y=fieldNumber)) + geom_point() + facet_wrap(~site)
check_missingdata_trapbymonth_wrapsite
check_missingdata_byfieldNumber <- ggplot(data, aes(x=monthName, y=fieldNumber)) + geom_point()
check_missingdata_byfieldNumber
check_missingdata_byfieldNumber_withcolor <- ggplot(data, aes(x=monthName, y=fieldNumber)) + geom_point(aes(color=site))
check_missingdata_byfieldNumber_withcolor
```

# Plot Halictidae family data over time by sex
```{r}
halictidaedata <- subset(data, family == "Halictidae")
halictidaedata <- subset(halictidaedata, sex == "MALE" | sex == "FEMALE")
halictidaebysex <- ggplot(halictidaedata, aes(x=month)) + geom_density(aes(color=sex))
halictidaebysex
halictidaebysex_occurence <- ggplot(halictidaedata, aes(x=monthName)) + geom_bar(aes(fill=sex), position=position_dodge())
halictidaebysex_occurence
```
# Plot specific groups within Halictidae family over time by sex
```{r}
agapostemondata <- subset(halictidaedata, genus == "Agapostemon")
agapostemonplot_bysex <- ggplot(agapostemondata, aes(x=monthName)) + 
  geom_bar(aes(fill=sex), position=position_dodge()) + ggtitle("Agapostemon")
agapostemonplot_bysex
agapostemonplot_bysex_bysite <- ggplot(agapostemondata, aes(x=monthName)) + 
  geom_bar(aes(fill=sex), position=position_dodge()) + ggtitle("Agapostemon") + facet_wrap(~site)
agapostemonplot_bysex_bysite
halictusdata <- subset(halictidaedata, genus == "Halictus")
halictusplot_bysex <- ggplot(halictusdata, aes(x=monthName)) + 
  geom_bar(aes(fill=sex), position=position_dodge()) + ggtitle("Halictus")
halictusplot_bysex
lasioglossumdata <- subset(halictidaedata, genus == "Lasioglossum")
lasioglossumplot_bysex <- ggplot(lasioglossumdata, aes(x=monthName)) + 
  geom_bar(aes(fill=sex), position=position_dodge()) + ggtitle("Lasioglossum")
lasioglossumplot_bysex
```
# Plot all bees over time by sex
```{r}
allbees_bysex <- ggplot(data, aes(x=monthName)) + geom_bar(aes(fill=sex), 
                                                               position=position_dodge())
allbees_bysex
allbees_bysex_density <- ggplot(data, aes(x=month)) + geom_density(aes(color=sex))
allbees_bysex_density
allbees_bysex_density_dayofyear <- ggplot(data, aes(x=startDayOfYear)) + geom_density(aes(color=sex))
allbees_bysex_density_dayofyear
```

# Plot native versus nonnative bees over time
```{r}
# Make new column for native or nonnative value
data$nativeStatus[data$species == "Apis mellifera"]<- "Nonnative"
data$nativeStatus[data$species != "Apis mellifera"]<- "Native"
nativestatusplot <- ggplot(data, aes(x=monthName)) +
  geom_bar(aes(fill=nativeStatus), position=position_dodge()) 
nativestatusplot
nativestatusplot_bysite <- ggplot(data, aes(x=monthName)) +
  geom_bar(aes(fill=nativeStatus), position=position_dodge()) + facet_wrap(~site)
nativestatusplot_bysite
nativestatusplot_density <- ggplot(data, aes(x=startDayOfYear)) + geom_density(aes(color=nativeStatus))
nativestatusplot_density
nativestatusplot_density_bysite <- ggplot(data, aes(x=startDayOfYear)) + geom_density(aes(color=nativeStatus)) + facet_wrap(~site)
nativestatusplot_density_bysite
```
# Overall Abundance
Plots of day of year versus overall abundance for all native bees, honey bees, Agapostemon bees, and Lasioglossum bees
```{r}
overallabundanceplot_nativebees <- ggplot(data, aes(monthName)) + geom_bar()
overallabundanceplot_nativebees
```


